### 2014-05-22 ms/ifm.tex
minor edits

\documentclass[12pt,aps,prb,preprint]{revtex4}
\usepackage{pgf}

\def\ket#1{|#1\rangle}
\def\X{{\bf X}}
\def\A{{\bf A}}

\begin{document}

\title{Quantum interaction-free measurement and a classical analog}

\author{Prasenjit Saha}
\email{psaha@physik.uzh.ch}
\affiliation{Physik-Institut, University of Zurich, Switzerland}
\author{Nick Kaiser}
\email{kaiser@hawaii.edu}
\affiliation{Institute for Astronomy, University of Hawaii at Manoa,
Honolulu, HI 96822}

\date{\today}

\begin{abstract}
Interaction-free measurement refers to quantum phenomena such as the
presence of a photon absorber being inferred without it absorbing any
photons.  We review the physics and its realization in interferometric
experiments.  We then consider a simple classical system: two coupled
oscillators, one of which is damped; the energy dissipation can be
made arbitrarily small by making the damping large enough.  The
phenomenon can be considered analogous to interaction-free
measurement.
\end{abstract}

\maketitle

\section{Introduction}

Many quantum paradoxes have been explored theoretically and
experimentally in recent years.  Most involve entangled states: the
famous EPR paradox is an old example, and quantum computing is a more
recent development\cite{2007qucosc.book}.  Entanglement is considered
an intrinsically quantum phenonemn classical analog.

There are also counter-intuitive quantum phenonema not involving
entanglement.  Imagine a photon, initially in a ``vertical'' polarized
state.  It is sent through a Faraday rotator, and then through a
vertically polarizing sheet.  The former rotates the polarization
plane towards ``horizontal'' by small angle, say $\pi/(2N)$, where $N$
is large.  The latter effectively measures the polarization state in
the vertical/horizontal basis; if the result comes out horizontal, the
photon is absorbed, otherwise it goes through in a vertical
eigenstate.  Since $N$ is large, the chances of being absorbed are
small.  Most likely, the photon will get slightly rotated and then
reset to vertical.  If the photon is re-routed back to the apparatus
$N$ times, it will still most likely emerge vertically polarized.  The
rotator and the polarizer appear to cancel each other out.  The
general phenomenon is known as the quantum Zeno
effect,\cite{PhysRevA.41.2295} but it can also be described in
everyday language, in terms of waking and sleeping
states.\cite{orzel2009teach}

What happens if we now remove the polarizer?  Clearly, the $N$ Faraday
rotations will take the photon to a horizontal state.  Thus, the net
effect of the polarizer is (most likely) to rotate a photon from
horizontal to vertical.  Recall that the polarizer is really just an
absorber of horizontal photons.  Yet it can, within a clever setup, be
made to rotate photons instead of absorbing them.  This is the essence
of interaction-free measurement.  Since the initial suggestion
(essentially the $N=1$ case)\cite{1993FoPh...23..987E} and
generalization to the now-standard form, together with the first
experiments\cite{PhysRevLett.74.4763} interaction-free measurement has
been the subject of many
demonstrations\cite{voorthuysen:1504,2006JPhB...39.3177N} and
reviews,\cite{deweerd:272} and applications aiming towards medical
imaging are starting to appear\cite{PhysRevA.80.040902}.  (An older
usage of similar
terminalogy\cite{springerlink:10.1007/BF01327019,dicke:925} refers to
a situation where a non-detection enables an inference.)

In this paper, we relate the above simple picture on the one hand to
the standard description of interaction-free measurement in terms of
an interferometric setup, and on the other hand to a classical system
of coupled oscillators.

\section{Quantum formulation}

The above thought experiment involving rotation the polarization plane
of photon is one simple way to approach the concept of
interaction-free measurement, but not the only one.  Another
well-known description, which more closely resembles some of the
actual experiments, is in terms of a Mach-Zehnder interferometer.  The
horizontal and vertical polarization states are replaced by the two
arms of the interferometer.  Figure~\ref{fig:mz} schematically shows
the $N=1$ case; the general case would be a concatenation of such
setups.

With these concrete descriptions in mind, let us formalize the process
of interaction-free measurement.  We write the two eigenstates in the
notation of cubits
\begin{equation}
\ket0 \equiv \left( \begin{array}{c} 1 \\ 0 \end{array} \right)  \qquad
\ket1 \equiv \left( \begin{array}{c} 0 \\ 1 \end{array} \right)
\end{equation}
and
\begin{equation}
\X \equiv \left( \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right)
\end{equation}
denotes a swap operator.  A beam splitter is given by the unitary operator
\begin{equation}
\exp\left[i\left({\textstyle\frac\pi2-\epsilon}\right)\right] \X \equiv
\sin\epsilon + i\cos\epsilon\,\X
\end{equation}
Then $\cos^2\epsilon$ is the reflectivity.  A mirror is simply $i\X$.

The consecutive operations of beam-splitting, reflection and beam
splitting can be written as
\begin{equation}
     (\sin\epsilon + i\cos\epsilon\,\X)
     \; i\X \;
     (\sin\epsilon + i\cos\epsilon\,\X)
\end{equation}
which simplifies to
\begin{equation}
     -(\sin2\epsilon + i\cos2\epsilon\,\X) \,.
\end{equation}
A cascade with beam-splitter followed by mirror $N-1$ times is
\begin{equation}
     \left[ (\sin\epsilon + i\cos\epsilon\,\X) \; i\X \right]^{N-1} \;
     (\sin\epsilon + i\cos\epsilon\,\X)
\end{equation}
which simplifies to
\begin{equation}
     (-1)^{N-1} (\sin N\epsilon + i\cos N\epsilon\,\X)  \,.
\end{equation}
If $N\epsilon=\pi/2$ this is identity, apart from a phase factor.

Now put an absorber in one of the arms.  Then we will have
\begin{equation}
     \left[ (\sin\epsilon + i\cos\epsilon\,\X) \; i\A \right]^{N-1} \;
     (\sin\epsilon + i\cos\epsilon\,\X)
  \qquad
  \A \equiv \left( \begin{array}{cc} 0 & 1 \\ 0 & 0 \end{array} \right)
\end{equation}
Working out the algebra gives
\begin{equation}
(-1)^{N-1} \, \cos^N\!\epsilon \,
\left( \begin{array}{cc} \tan\epsilon & -i\tan^2\epsilon \\
                         i            & \tan\epsilon 
       \end{array} \right)
\end{equation}
If $\epsilon$ is small (very slightly transmissive) this operator
changes $\ket0$ to $\ket1$.

\section{A classical analogy}

Consider two coupled oscillators, one of them damped.
\begin{equation}
\begin{array}{l}
\ddot x + x + \kappa y = 0 \\
\ddot y + y + \kappa x + \gamma y = 0
\end{array}
\end{equation}
If $\gamma=0$ has solutions of the form
\begin{equation}
\begin{array}{l}
x = \cos(\sqrt{1+\kappa}\,t) + \cos(\sqrt{1-\kappa}\,t)  \\
y = \cos(\sqrt{1+\kappa}\,t) - \cos(\sqrt{1-\kappa}\,t)
\end{array}
\end{equation}
and the amplitude sloshes from one to the other.

With damping, easiest to solve using Laplace transform. For the
initial conditions $x(0)=1$ and $\dot x(0) = y(0) = \dot y(0) = 0$ we
have
\begin{equation}
\begin{array}{l}
s^2 X - s + X + \kappa Y = 0 \\
s^2 Y + Y + \kappa X + \gamma s Y = 0
\end{array}
\end{equation}
For strong damping, we have to $O(1/\gamma)$
\begin{equation}
\begin{array}{l}
\displaystyle
X = \frac s{s^2+1} + \frac{\kappa^2}\gamma \frac1{(s^2+1)^2} \\
\displaystyle
Y = -\frac\kappa\gamma \frac1{s^2+1}
\end{array}
\end{equation}
Solution to $O(1/\gamma)$
\begin{equation}
\begin{array}{l}
\displaystyle
x = \cos t + \frac{\kappa^2}{2\gamma} (\sin t - t\cos t) \\
\displaystyle
y = -\frac\kappa\gamma \sin t
\end{array}
\end{equation}
Amplitude damping rate $\propto\kappa^2/\gamma$.

For mechanical oscillators, it may be counter-intuitive that more
damping means less energy dissipation.  In the electrical case (an LC
circuit coupled to an RLC circuit) it is obvious that increasing
resistance reduces dissipation.

\bibliography{ifm.bib}

\newpage

\def\vmargin{\hrule height 1cm width 0pt}

\def\pgfbeamsplitter(#1,#2)#3{\pgfputat{\pgfxy(#1,#2)}
                              {\pgfbox[center,center]#3
                               \pgfxyline(-1.1,0)(0,1.1)
                               \pgfxyline(0,1.1)(1.1,0)
                               \pgfxyline(1.1,0)(0,-1.1)
                               \pgfxyline(0,-1.1)(-1.1,0)}}

\def\pgfbeamsplitter(#1,#2)#3{\pgfputat{\pgfxy(#1,#2)}
                              {\pgfbox[center,center]#3
                               \pgfxyline(-1.1,0)(0,1.1)
                               \pgfxyline(0,1.1)(1.1,0)
                               \pgfxyline(1.1,0)(0,-1.1)
                               \pgfxyline(0,-1.1)(-1.1,0)}}

\def\semireflecting{$\hbox{beam}\atop\hbox{splitter}$}
\def\arbireflecting{$e^{i\alpha\X}$}
\def\nearlytransparent{$e^{i\left(\frac\pi2-\epsilon\right)\X}$}

\def\pgfsimpmirror(#1,#2){\pgfputat{\pgfxy(#1,#2)}
                      {\pgfbox[center,center]{M}
                      \pgfxyline(-.4,-.4)(.4,-.4)
                      \pgfxyline(.4,-.4)(.4,.4)
                      \pgfxyline(.4,.4)(-.4,.4)
                      \pgfxyline(-.4,.4)(-.4,-.4)}}

\def\pgfmirror(#1,#2){\pgfputat{\pgfxy(#1,#2)}
                      {\pgfbox[center,center]{$i\X$}
                      \pgfxyline(-.4,-.4)(.4,-.4)
                      \pgfxyline(.4,-.4)(.4,.4)
                      \pgfxyline(.4,.4)(-.4,.4)
                      \pgfxyline(-.4,.4)(-.4,-.4)}}
\def\pgfabsorber(#1,#2){\pgfputat{\pgfxy(#1,#2)}
                      {\pgfbox[center,center]{$\bf 0$}
                      \pgfxyline(-.4,-.4)(.4,-.4)
                      \pgfxyline(.4,-.4)(.4,.4)
                      \pgfxyline(.4,.4)(-.4,.4)
                      \pgfxyline(-.4,.4)(-.4,-.4)}}

\begin{figure}
\begin{center}
\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfsimpmirror(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5) 
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfsimpmirror(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5, 1.5)(-.7, .7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,-.7)(5.5,-1.5) 
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\semireflecting
\pgfabsorber(2,2) 
\pgfsimpmirror(2,-2)
\pgfbeamsplitter(4,0)\semireflecting
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5) 
\pgfxyline(4.7,-.7)(5.5,-1.5) 
\end{pgfpicture}
\end{center}
\caption{\label{fig:mz} Schematic representation of interaction-free
  measurement. Each diagram stands for a Mach-Zehnder set up so that
  both arms are exactly equal.  Accordingly, a photon sent axially
  into one arm emerges parallel to the original direction.  This is
  seen in the upper and middle diagrams.  In the lower diagram, one of
  the mirrors is replaced by an absorber, and as a result, photons can
  emerge perpendicular to their original direction.  Such photons
  effectively detect the absorber without getting absorbed.}
\end{figure}

\begin{figure}
\begin{center}
\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfmirror(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7,-1.7)}{\pgfbox[right,center]{$\ket0$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-\sin2\epsilon\;\ket0$}}
         \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-i\cos2\epsilon\;\ket1$}}
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfmirror(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5, 1.5)(-.7, .7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,1.5)(3.3,.7) 
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7, 1.7)}{\pgfbox[right,center]{$\ket1$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-i\cos2\epsilon\;\ket0$}}
          \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-\sin2\epsilon\;\ket1$}}
\end{pgfpicture}
\vmargin\vmargin\vmargin
\begin{pgfpicture}
\pgfbeamsplitter(0,0)\nearlytransparent 
\pgfabsorber(2,2) 
\pgfmirror(2,-2)
\pgfbeamsplitter(4,0)\nearlytransparent 
\pgfsetendarrow{\pgfarrowsingle}
\pgfxyline(-1.5,-1.5)(-.7,-.7) 
\pgfxyline(.7,.7)(1.5,1.5)
\pgfxyline(.7,-.7)(1.5,-1.5)
\pgfxyline(2.5,-1.5)(3.3,-.7)
\pgfxyline(4.7,.7)(5.5,1.5)
\pgfxyline(4.7,-.7)(5.5,-1.5)
\pgfputat{\pgfxy(-1.7,-1.7)}{\pgfbox[right,center]{$\ket0$}} 
\pgfputat{\pgfxy(5.7, 1.7)}{\pgfbox[left,center]
         {$-\sin\epsilon\cos\epsilon\;\ket0$}}
          \pgfputat{\pgfxy(5.7,-1.7)}{\pgfbox[left,center]
         {$-i\cos^2\epsilon\;\ket1$}}
\end{pgfpicture}
\end{center}
\caption{\label{fig:qubits} Like Figure \ref{fig:mz} but annotated
  with quantum states and operators.  A photon is denoted by $\ket0$
  if moving towards upper right, and by $\ket1$ if moving towards
  lower right.  Beam splitters and mirrors are unitary operators.  An
  absorber is a non-unitary operator, which zeros the $\ket0$
  subspace.}
\end{figure}

\end{document}

### 2015-07-17 ms/ifm.tex
editing

@@
 
-\title{Quantum interaction-free measurement and a classical analog}
+\title{A Classical Analog of Quantum Interaction-Free Measurement}
 
@@
 \begin{abstract}
-Interaction-free measurement refers to quantum phenomena such as the
-presence of a photon absorber being inferred without it absorbing any
-photons.  We review the physics and its realization in interferometric
-experiments.  We then consider a simple classical system: two coupled
-oscillators, one of which is damped; the energy dissipation can be
-made arbitrarily small by making the damping large enough.  The
-phenomenon can be considered analogous to interaction-free
-measurement.
+Interaction-free measurement is usually discussed in quantum language and
+refers to phenomena such as the presence of a photon absorber being
+inferred without it actually absorbing any photons.  Here we make the
+simple observation that there is an
+exactly analogous situation in classical mechanics: If one has
+a pair of identical coupled oscillators A \& B, and A is set in motion,
+then one will see the strength of the oscillation of A slowly decrease
+with time, and over longer times the energy will periodically `slosh' back
+and forth between the two systems.  But if instead oscillator B is damped -- so
+it is able to {\em absorb\/} energy -- the rate of decrease of the
+energy in oscillator A will be suppressed.   
+Thus one can infer the ability of oscillator B to 
+absorb energy by observing the behaviour of oscillator A without, in
+the limit that the damping is strong, there actually being any
+energy absorbed.
+% We review the physics and its realization in interferometric
+% experiments.  We then consider a simple classical system: two coupled
+% oscillators, one of which is damped; the energy dissipation can be
+% made arbitrarily small by making the damping large enough.  The
+% phenomenon can be considered analogous to interaction-free
+% measurement.
 \end{abstract}
@@
 recent development\cite{2007qucosc.book}.  Entanglement is considered
-an intrinsically quantum phenonemn classical analog.
+an intrinsically quantum phenonemon with no classical analog.
 
@@
 vertically polarizing sheet.  The former rotates the polarization
-plane towards ``horizontal'' by small angle, say $\pi/(2N)$, where $N$
+plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where $N$
 is large.  The latter effectively measures the polarization state in
@@
 eigenstate.  Since $N$ is large, the chances of being absorbed are
-small.  Most likely, the photon will get slightly rotated and then
+small.  More specifically, after the rotation the {\em amplitude\/}
+to be in the horizontal state will be of order $1/N$ so the probability to be
+in that mode -- and therefore the probability of absorption -- is of order $1/N^2$.
+Thus, most likely, the photon will get slightly rotated and then
 reset to vertical.  If the photon is re-routed back to the apparatus
@@
 What happens if we now remove the polarizer?  Clearly, the $N$ Faraday
-rotations will take the photon to a horizontal state.  Thus, the net
-effect of the polarizer is (most likely) to rotate a photon from
-horizontal to vertical.  Recall that the polarizer is really just an
-absorber of horizontal photons.  Yet it can, within a clever setup, be
-made to rotate photons instead of absorbing them.  This is the essence
-of interaction-free measurement.  Since the initial suggestion
+rotations will then convert the photon to a horizontal state.  Thus, the net
+effect of the polarizer is to rotate a photon from
+horizontal to vertical.  Despite the fact that the polarizer is really just an
+absorber of horizontal photons, it can, within this experimental set-up, have
+the effect of rotating photons instead of absorbing them.  This is the essence
+of interaction-free measurement; we can infer, or `measure' the presence or absence
+of the polarizer by measuring the polarization.  If the polarization is rotated
+we know that the absorber cannot have been present -- and there is then of course
+no absorption -- while if the polarization is not changed then we know that
+the absorber was present, but again, at least in the limit $N \rightarrow \infty$,
+no absorption actually takes place.
+
+Since the initial suggestion
 (essentially the $N=1$ case)\cite{1993FoPh...23..987E} and
@@
 usage of similar
-terminalogy\cite{springerlink:10.1007/BF01327019,dicke:925} refers to
+terminology\cite{springerlink:10.1007/BF01327019,dicke:925} refers to
 a situation where a non-detection enables an inference.)
@@
 
-\section{Quantum formulation}
+\section{Quantum/Wave-Mechanical formulation}
 
@@
 of interaction-free measurement.  We write the two eigenstates in the
-notation of cubits
+notation of qbits
 \begin{equation}
@@
 
-\section{A classical analogy}
+The discussion above uses the quantum mechanical language of
+`states', but the behaviour of such an interferometer can, of course,
+but understood entirely as a purely classical wave obeying
+Maxwell's equations.
 
-Consider two coupled oscillators, one of them damped.
+\section{The classical analogy}
+
+To flesh out the observation of the abstract, 
+consider two coupled oscillators, one of them damped.
 \begin{equation}
@@
 \end{equation}
-If $\gamma=0$ has solutions of the form
+In the absence of damping ($\gamma=0$), it is easy to verify that
 \begin{equation}
 \begin{array}{l}
-x = \cos(\sqrt{1+\kappa}\,t) + \cos(\sqrt{1-\kappa}\,t)  \\
-y = \cos(\sqrt{1+\kappa}\,t) - \cos(\sqrt{1-\kappa}\,t)
+2x = \cos(\sqrt{1+\kappa}\,t) + \cos(\sqrt{1-\kappa}\,t)  \\
+2y = \cos(\sqrt{1+\kappa}\,t) - \cos(\sqrt{1-\kappa}\,t)
 \end{array}
 \end{equation}
-and the amplitude sloshes from one to the other.
+is a solution for the initial conditions
+\begin{equation}
+x(0) = 1, y(0) = \dot x(0) = \dot y(0) = 0 \,.
+\end{equation}
+If the coupling is weak ($\kappa\ll1$) the solution takes the form
+\begin{equation}
+\begin{array}{l}
+x = \cos(\frac12\kappa t) \cos t \\
+y = -\sin(\frac12\kappa t) \sin t \\
+\end{array}
+\end{equation}
+and the amplitude sloshes from $x$ to $y$.  The weaker the coupling,
+the slower the transfer of energy between the oscillators.
 
@@
   measurement. Each diagram stands for a Mach-Zehnder set up so that
-  both arms are exactly equal.  Accordingly, a photon sent axially
-  into one arm emerges parallel to the original direction.  This is
-  seen in the upper and middle diagrams.  In the lower diagram, one of
-  the mirrors is replaced by an absorber, and as a result, photons can
-  emerge perpendicular to their original direction.  Such photons
-  effectively detect the absorber without getting absorbed.}
+  both arms are exactly equal.  In the upper and middle diagrams, a
+  photon sent axially into one arm emerges parallel to the original
+  direction.  In the lower diagram, one of the mirrors is replaced by
+  an absorber, and as a result, photons can emerge perpendicular to
+  their original direction.  Such photons effectively detect the
+  absorber without getting absorbed.}
 \end{figure}
@@
 \caption{\label{fig:qubits} Like Figure \ref{fig:mz} but annotated
-  with quantum states and operators.  A photon is denoted by $\ket0$
-  if moving towards upper right, and by $\ket1$ if moving towards
-  lower right.  Beam splitters and mirrors are unitary operators.  An
-  absorber is a non-unitary operator, which zeros the $\ket0$
-  subspace.}
+  with states and operators.  A photon is denoted by $\ket0$ if moving
+  towards upper right, and by $\ket1$ if moving towards lower right.
+  Beam splitters and mirrors are unitary operators.  An absorber is a
+  non-unitary operator, which zeros the $\ket0$ subspace.}
 \end{figure}

### 2015-07-17 ms/ifm.tex
editing

@@
 vertically polarizing sheet.  The former rotates the polarization
-plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where $N$
-is large.  The latter effectively measures the polarization state in
-the vertical/horizontal basis; if the result comes out horizontal, the
-photon is absorbed, otherwise it goes through in a vertical
+plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where
+$N$ is large.  The latter effectively measures the polarization state
+in the vertical/horizontal basis; if the result comes out horizontal,
+the photon is absorbed, otherwise it goes through in a vertical
 eigenstate.  Since $N$ is large, the chances of being absorbed are
-small.  More specifically, after the rotation the {\em amplitude\/}
-to be in the horizontal state will be of order $1/N$ so the probability to be
-in that mode -- and therefore the probability of absorption -- is of order $1/N^2$.
-Thus, most likely, the photon will get slightly rotated and then
-reset to vertical.  If the photon is re-routed back to the apparatus
-$N$ times, it will still most likely emerge vertically polarized.  The
-rotator and the polarizer appear to cancel each other out.  The
-general phenomenon is known as the quantum Zeno
+small.  More specifically, after the rotation the {\em amplitude\/} to
+be in the horizontal state will be $\sim1/N$ so the probability to be
+in that mode --- and therefore the probability of absorption --- will
+be $\sim1/N^2$.  Thus, most likely, the photon will get slightly
+rotated and then reset to vertical.  If the photon is re-routed back
+to the apparatus $N$ times, it will still most likely emerge
+vertically polarized.  The rotator and the polarizer appear to cancel
+each other out.  The general phenomenon is known as the quantum Zeno
 effect,\cite{PhysRevA.41.2295} but it can also be described in
@@
 What happens if we now remove the polarizer?  Clearly, the $N$ Faraday
-rotations will then convert the photon to a horizontal state.  Thus, the net
-effect of the polarizer is to rotate a photon from
-horizontal to vertical.  Despite the fact that the polarizer is really just an
-absorber of horizontal photons, it can, within this experimental set-up, have
-the effect of rotating photons instead of absorbing them.  This is the essence
-of interaction-free measurement; we can infer, or `measure' the presence or absence
-of the polarizer by measuring the polarization.  If the polarization is rotated
-we know that the absorber cannot have been present -- and there is then of course
-no absorption -- while if the polarization is not changed then we know that
-the absorber was present, but again, at least in the limit $N \rightarrow \infty$,
-no absorption actually takes place.
+rotations will then convert the photon to a horizontal state.  Thus,
+the net effect of the polarizer is to rotate a photon from horizontal
+to vertical.  Despite the fact that the polarizer is really just an
+absorber of horizontal photons, it can, within this experimental
+set-up, have the effect of rotating photons instead of absorbing them.
+This is the essence of interaction-free measurement; we can infer, or
+`measure' the presence or absence of the polarizer by measuring the
+polarization.  If the polarization is rotated we know that the
+absorber cannot have been present -- and there is then of course no
+absorption -- while if the polarization is not changed then we know
+that the absorber was present, but again, at least in the limit $N
+\rightarrow \infty$, no absorption actually takes place.
 
-Since the initial suggestion
-(essentially the $N=1$ case)\cite{1993FoPh...23..987E} and
-generalization to the now-standard form, together with the first
-experiments\cite{PhysRevLett.74.4763} interaction-free measurement has
-been the subject of many
-demonstrations\cite{voorthuysen:1504,2006JPhB...39.3177N} and
-reviews,\cite{deweerd:272} and applications aiming towards medical
-imaging are starting to appear\cite{PhysRevA.80.040902}.  (An older
-usage of similar
+Since the initial suggestion (essentially the $N=1$
+case)\cite{1993FoPh...23..987E} and generalization to the now-standard
+form, together with the first experiments\cite{PhysRevLett.74.4763}
+interaction-free measurement has been the subject of many
+demonstrations\cite{PhysRevA.57.3987,voorthuysen:1504,2006JPhB...39.3177N}
+and reviews.\cite{deweerd:272} (An older usage of similar
 terminology\cite{springerlink:10.1007/BF01327019,dicke:925} refers to
-a situation where a non-detection enables an inference.)
+a situation where a non-detection enables an inference.)  Applications
+to non-invasive medical imaging\cite{PhysRevA.80.040902} are
+particularly interesting.
 
@@
 of coupled oscillators.
-
-\section{Quantum/Wave-Mechanical formulation}
 
@@
 the $N=1$ case; the general case would be a concatenation of such
-setups.
+devices.
 
-With these concrete descriptions in mind, let us formalize the process
-of interaction-free measurement.  We write the two eigenstates in the
-notation of qbits
+\section{An interferometer as an absorber detector}
+
+Let us express the interferometric setup of Figure~\ref{fig:mz} in the
+language of quantum two-state systems, borrowing the notation of
+quantum computing.  Figure~\ref{fig:qubits} replaces the optical
+elements with states and operators.  The two eigenstates are rays
+moving towards upper right, or towards lower right in the figures. We
+write these states as kets or matrices as
 \begin{equation}
@@
 \end{equation}
-and
+while
 \begin{equation}
@@
 \end{equation}
-denotes a swap operator.  A beam splitter is given by the unitary operator
+denotes an operator swapping these states.  A beam splitter is given
+by the unitary operator
 \begin{equation}
@@
 \end{equation}
-Then $\cos^2\epsilon$ is the reflectivity.  A mirror is simply $i\X$.
+Then $\cos^2\epsilon$ is the reflectivity.  Small $\epsilon$ means
+highly reflective, and $\epsilon=0$ corresponds to a mirror, whose
+operator is $i\X$.  We emphasize, however, that the quantum notation
+does not mean the interferometer is intrinsically quantum.  The
+behavior of such an interferometer can also be understood in terms of
+classical waves obeying Maxwell's equations.
 
-The consecutive operations of beam-splitting, reflection and beam
+Consecutive operations of beam-splitting, reflection and beam
 splitting can be written as
@@
 \begin{equation}
-     -(\sin2\epsilon + i\cos2\epsilon\,\X) \,.
+     -(\sin2\epsilon + i\cos2\epsilon\,\X)
+\label{eqn:no-absorber}
 \end{equation}
-A cascade with beam-splitter followed by mirror $N-1$ times is
+and is illustrated by the upper and middle diagrams in
+Figure~\ref{fig:qubits}.
+
+If we now add more elements to the interferometer, specifically $N-1$
+sets of beam-splitters alternating with mirrors, the resulting state
+is
 \begin{equation}
@@
 
-Now put an absorber in one of the arms.  Then we will have
+Now let us put an absorber in one of the arms.  This blocks the
+$\ket0$ state, while the mirror (in the other arm) reflects $\ket1$ as
+before.  Then we will have
 \begin{equation}
@@
 \end{equation}
-Working out the algebra gives
+involving a non-Hermitian operator $\A$.  Working out the algebra
+gives
 \begin{equation}
@@
        \end{array} \right)
+\label{eqn:absorber}
 \end{equation}
 If $\epsilon$ is small (very slightly transmissive) this operator
-changes $\ket0$ to $\ket1$.
+changes $\ket0$ to $\ket1$.  The lower diagram in
+Figure~\ref{fig:qubits} illustrates the result for $N=1$.
 
-The discussion above uses the quantum mechanical language of
-`states', but the behaviour of such an interferometer can, of course,
-but understood entirely as a purely classical wave obeying
-Maxwell's equations.
+The contrasting results (\ref{eqn:no-absorber}) and
+(\ref{eqn:absorber}) express the characteristic feature of
+interaction-free measurement.  With no absorber the amplitude cycles
+between $\ket0$ and $\ket1$.  An absorber blocks the sloshing, without
+necessarily absorbing.
 
-\section{The classical analogy}
+\section{Coupled oscillators as an absorber detector}
 
-To flesh out the observation of the abstract, 
-consider two coupled oscillators, one of them damped.
+In the previous section we considered a system that could be quantum,
+or classically wave-optic.  In this section, we show that the same
+qualitative behavior emerges in a purely classical system.
+
+Consider two coupled oscillators, one of them damped.  Such a system
+can be written, with a suitable choice of units, as
 \begin{equation}
@@
 \ddot x + x + \kappa y = 0 \\
-\ddot y + y + \kappa x + \gamma y = 0
+\ddot y + y + \kappa x + \gamma y = 0 \,.
+\label{eqn:coupled}
 \end{array}
 \end{equation}
-In the absence of damping ($\gamma=0$), it is easy to verify that
+Without the damping term ($\gamma=0$), it is easy to verify that
 \begin{equation}
@@
 x(0) = 1, y(0) = \dot x(0) = \dot y(0) = 0 \,.
+\label{eqn:inic}
 \end{equation}
@@
 \begin{array}{l}
-x = \cos(\frac12\kappa t) \cos t \\
-y = -\sin(\frac12\kappa t) \sin t \\
+x = \cos\left(\frac12\kappa t\right) \cos t \\
+y = -\sin\left(\frac12\kappa t\right) \sin t \\
 \end{array}
 \end{equation}
-and the amplitude sloshes from $x$ to $y$.  The weaker the coupling,
-the slower the transfer of energy between the oscillators.
+and the amplitude sloshes from $x$ to $y$.
 
-With damping, easiest to solve using Laplace transform. For the
-initial conditions $x(0)=1$ and $\dot x(0) = y(0) = \dot y(0) = 0$ we
-have
+Now putting back the damping, one can solve the equations in different
+ways.  A concise way is to use Laplace transforms.  Taking the Laplace
+transform of the equations (\ref{eqn:coupled}) with the same initial
+conditions (\ref{eqn:inic}) gives
 \begin{equation}
@@
 s^2 X - s + X + \kappa Y = 0 \\
-s^2 Y + Y + \kappa X + \gamma s Y = 0
+s^2 Y + Y + \kappa X + \gamma s Y = 0 \,.
 \end{array}
@@
 \end{equation}
-Solution to $O(1/\gamma)$
+Taking the inverse Laplace transform and simplifying gives the
+solution to $O(1/\gamma)$
 \begin{equation}
@@
 \displaystyle
-y = -\frac\kappa\gamma \sin t
+y = -\frac\kappa\gamma \sin t \,.
 \end{array}
 \end{equation}
-Amplitude damping rate $\propto\kappa^2/\gamma$.
+As we see, damping slows down the transfer to $y$.
 
-For mechanical oscillators, it may be counter-intuitive that more
-damping means less energy dissipation.  In the electrical case (an LC
-circuit coupled to an RLC circuit) it is obvious that increasing
-resistance reduces dissipation.
+In a mechanical system, it may seem odd that more damping means less
+energy dissipation.  But for electrical oscillotors (an LC circuit
+coupled to an RLC circuit) we are used to the idea that increasing
+resistance reduces dissipation.  Quantum interaction-free measurement
+can thus be seen simply as a quantized version of such a system.
 

### 2015-07-17 ms/ifm.tex
editing

@@
 \begin{abstract}
-Interaction-free measurement is usually discussed in quantum language and
-refers to phenomena such as the presence of a photon absorber being
-inferred without it actually absorbing any photons.  Here we make the
-simple observation that there is an
-exactly analogous situation in classical mechanics: If one has
-a pair of identical coupled oscillators A \& B, and A is set in motion,
-then one will see the strength of the oscillation of A slowly decrease
-with time, and over longer times the energy will periodically `slosh' back
-and forth between the two systems.  But if instead oscillator B is damped -- so
-it is able to {\em absorb\/} energy -- the rate of decrease of the
-energy in oscillator A will be suppressed.   
-Thus one can infer the ability of oscillator B to 
-absorb energy by observing the behaviour of oscillator A without, in
-the limit that the damping is strong, there actually being any
-energy absorbed.
-% We review the physics and its realization in interferometric
-% experiments.  We then consider a simple classical system: two coupled
-% oscillators, one of which is damped; the energy dissipation can be
-% made arbitrarily small by making the damping large enough.  The
-% phenomenon can be considered analogous to interaction-free
-% measurement.
+Interaction-free measurement generally refers to quantum phenomena
+such as the presence of a photon absorber being inferred without it
+actually absorbing any photons.  Here we make the simple observation
+that there is an analogous situation in classical mechanics: two
+coupled oscillators, one of which is damped; the energy dissipation
+can be made arbitrarily small by making the damping large enough, thus
+revealing the presence of the damper despite arbitrarily low energy
+dissipation.
 \end{abstract}
@@
 
-In this paper, we relate the above simple picture on the one hand to
-the standard description of interaction-free measurement in terms of
-an interferometric setup, and on the other hand to a classical system
-of coupled oscillators.
+In this paper, we argue that interaction-free measurement need not be
+quantum at all.  Rather, the characteristics features of (a)~an
+amplitude that cycles between two states, but (b)~an absorber can
+block this cycling, thus revealing its presence, but without actually
+absorbing any energy, can also appear in classical systems.  If one
+has a pair of identical coupled oscillators $x$ and $y$, and $x$ is
+set in motion, one will see the strength of the oscillation of $x$
+slowly decrease with time, and over longer times the energy will
+periodically `slosh' back and forth between the two systems.  But if
+instead oscillator $y$ is damped --- so it is able to {\em absorb\/}
+energy --- the rate of decrease of the energy in oscillator $x$ will
+be suppressed.  Thus one can infer the ability of oscillator $y$ to
+absorb energy by observing the behaviour of oscillator $x$ without, in
+the limit that the damping is strong, there actually being any energy
+absorbed.
 

### 2015-07-17 ms/ifm.tex
editing

@@
 each other out.  The general phenomenon is known as the quantum Zeno
-effect,\cite{PhysRevA.41.2295} but it can also be described in
-everyday language, in terms of waking and sleeping
+effect,\cite{1977JMP....18..756M,PhysRevA.41.2295} but it can also be
+described in everyday language, in terms of waking and sleeping
 states.\cite{orzel2009teach}
@@
 slowly decrease with time, and over longer times the energy will
-periodically `slosh' back and forth between the two systems.  But if
+periodically slosh back and forth between the two systems.  But if
 instead oscillator $y$ is damped --- so it is able to {\em absorb\/}
@@
 
-The above thought experiment involving rotation the polarization plane
-of photon is one simple way to approach the concept of
-interaction-free measurement, but not the only one.  Another
-well-known description, which more closely resembles some of the
-actual experiments, is in terms of a Mach-Zehnder interferometer.  The
-horizontal and vertical polarization states are replaced by the two
-arms of the interferometer.  Figure~\ref{fig:mz} schematically shows
-the $N=1$ case; the general case would be a concatenation of such
-devices.
+In the other sections of this paper, we flesh out the above argument.
+In section~\ref{sec:interf} we work through an alternative well-known
+description of interaction-free measurement, more closely resembles
+some of the actual experiments.  This is in terms of a Mach-Zehnder
+interferometer, whose two arms replace the horizontal and vertical
+polarization states of the above thought-experiment.
+Figure~\ref{fig:mz} schematically shows the $N=1$ case; the general
+case would be a concatenation of such devices.  With an absorber, the
+amplitude sloshes between the 'lower' and 'upper' arms of the
+interferometer; the absorber results in confining the amplitude to the
+lower arm.  Section~\ref{sec:oscill} briefly solves for the
+oscillators $x$ and $y$ from the previous paragraph.
 
-\section{An interferometer as an absorber detector}
+\section{An interferometer as an absorber detector}\label{sec:interf}
 
@@
 
-\section{Coupled oscillators as an absorber detector}
+\section{Coupled oscillators as an absorber detector}\label{sec:oscill}
 

### 2015-07-17 ms/ifm.tex
ready to submit?

@@
-\documentclass[12pt,aps,prb,preprint]{revtex4}
+\documentclass[prb,preprint]{revtex4-1}
 \usepackage{pgf}
@@
 There are also counter-intuitive quantum phenonema not involving
-entanglement.  Imagine a photon, initially in a ``vertical'' polarized
+entanglement.  Imagine a photon, initially in a `vertical' polarized
 state.  It is sent through a Faraday rotator, and then through a
 vertically polarizing sheet.  The former rotates the polarization
-plane towards ``horizontal'' by a small angle, say $\pi/(2N)$, where
+plane towards `horizontal' by a small angle, say $\pi/(2N)$, where
 $N$ is large.  The latter effectively measures the polarization state
@@
 This is the essence of interaction-free measurement; we can infer, or
-`measure' the presence or absence of the polarizer by measuring the
+`measure', the presence or absence of the polarizer by measuring the
 polarization.  If the polarization is rotated we know that the
@@
 In this paper, we argue that interaction-free measurement need not be
-quantum at all.  Rather, the characteristics features of (a)~an
+quantum at all.  Rather, the characteristic features of (a)~an
 amplitude that cycles between two states, but (b)~an absorber can
@@
 case would be a concatenation of such devices.  With an absorber, the
-amplitude sloshes between the 'lower' and 'upper' arms of the
+amplitude sloshes between the `lower' and `upper' arms of the
 interferometer; the absorber results in confining the amplitude to the
-lower arm.  Section~\ref{sec:oscill} briefly solves for the
-oscillators $x$ and $y$ from the previous paragraph.
+lower arm.  In section~\ref{sec:oscill} we validate the foregoing
+argument by briefly solving for the coupled oscillators $x$ and $y$
+from the previous paragraph.
 
-\section{An interferometer as an absorber detector}\label{sec:interf}
+\section{Interferometers as detectors of absorbers}\label{sec:interf}
 
@@
 (\ref{eqn:absorber}) express the characteristic feature of
-interaction-free measurement.  With no absorber the amplitude cycles
-between $\ket0$ and $\ket1$.  An absorber blocks the sloshing, without
-necessarily absorbing.
+interaction-free measurement.  With no absorber, the amplitude out of
+the $N$-th beam splitter sloshes between $\ket0$ and $\ket1$ as $N$
+increases.  An absorber blocks the sloshing, without necessarily
+absorbing.
 
-\section{Coupled oscillators as an absorber detector}\label{sec:oscill}
+\section{Coupled oscillators as detectors of absorbers}\label{sec:oscill}
 
@@
 
-Consider two coupled oscillators, one of them damped.  Such a system
-can be written, with a suitable choice of units, as
+Consider two coupled oscillators, $x,y$, of which $y$ is damped.  Such
+a system can be written, with a suitable choice of units, as
 \begin{equation}
